---
title: "Geography 176A"
author: '[Liam Finn](https://anthonyfinn.github.io/liam.finn/)'
subtitle: 'Lab 03: Distances and the Border Zone'
output:
  html_document:
    theme: darkly
---

```{r}
library(tidyverse)
library(sf)
library(units)
library(ggplot2)
library(ggrepel)
library(gghighlight)
library(ggthemes)
library(knitr)
```

```{r}
eqdc = '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```


```{r}

CONUS= USAboundaries::us_states(resolution = 'low') %>% 
  filter(!name %in% c("Puerto Rico", "Alaska", "Hawaii", "District of Columbia"))

CONUS<- st_transform(CONUS, eqdc)
st_crs(CONUS)
plot(CONUS$geometry)

MEX= rnaturalearth::countries110 %>% 
  st_as_sf() %>% 
  filter(admin== 'Mexico')
plot(MEX$geometry)

CAN= rnaturalearth::countries110 %>% 
  st_as_sf() %>% 
  filter(admin== 'Canada')
plot(CAN$geometry)
```




QUESTION #1


1.2 Get US State boundaries
```{r}
region = data.frame(region = state.region, 
                    state_name = state.name)
USA = right_join(USAboundaries::us_states(),
                   region,
                   by = "state_name") %>% 
  filter(!state_abbr %in% c("HI", "PR", "AK")) 

USA <- st_transform(USA, '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs')
  
plot(USA$geometry, main = "Continental United States ")
```

1.3 - Get country boundaries for Mexico, the United States of America, and Canada
```{r}
library(rnaturalearth)

boundaries = rnaturalearth::countries110 %>% 
  st_as_sf(crs = 4269) %>% 
  filter(sovereignt%in% c ("United States of America", "Canada", "Mexico")) %>% 
  st_transform(eqdc)

plot(boundaries$geometry)
```

1.4 - Get city locations

```{r}
cities = readr::read_csv('/Users/lfinn443/github/geog-176A-labs/data/uscities.csv') %>%
  st_as_sf(coords = c("lng", "lat"), crs = 4326) %>%  
  st_transform(eqdc) %>% 
  st_filter(CONUS)
```


QUESTION 2.1
```{r}
USA_u = st_union(USA) %>%
  st_cast("MULTILINESTRING") %>% 
  st_transform(st_crs(cities))
plot(USA_u)

#DISTANCE TO US BORDER
cities = cities %>% 
  mutate(dist_to_border = st_distance(cities, USA_u),
         dist_to_border = units::set_units(dist_to_border, "km"),
         dist_to_border = units::drop_units(dist_to_border))

farthest_cities = cities %>%
  slice_max(dist_to_border, n = 5) %>% 
  select(city, state_name, dist_to_border) %>% 
  st_drop_geometry()

kable(farthest_cities, caption = "Furthest US Cities from National Border",
      col.names = c("City", "State", "Nearest US Border/Coastline: km"),
      format.args = list(big.mark = ","))
```

QUESTION 2.2
```{r}
USA_c=st_combine(USA)  %>%
  st_cast("MULTILINESTRING") %>% 
  st_transform(st_crs(cities))
plot(USA_c)

#Distance to States
cities = cities %>% 
  mutate(dist_to_state = st_distance(cities, USA_c),
         dist_to_state= units::set_units(dist_to_state, "km"),
         dist_to_state = units::drop_units(dist_to_state))
cities_far_state=cities %>%
  slice_max(dist_to_state, n = 5) %>% 
  select(city, state_name, dist_to_state) %>% 
  st_drop_geometry()

kable(cities_far_state, caption = "Furthest US Cities from State Border",
      col.names = c("City", "State", "Distance to State Border: km"),
      format.args = list(big.mark = ","))
  
```
QUESTION 2.3 - Distance to Mexico (km)
```{r}
MEX= boundaries %>% 
  filter(sovereignt == "Mexico") %>% 
  st_transform(st_crs(cities))

cities = cities %>% 
  mutate(MEX_dist = st_distance(cities, MEX),
         MEX_dist = units::set_units(MEX_dist, "km"),
         MEX_dist = units::drop_units(MEX_dist))

#furthest US Cities from MEX
 UScities_MEX= cities %>% 
  slice_max(MEX_dist, n = 5) %>% 
  select(city, state_name, MEX_dist) %>% 
  st_drop_geometry()

kable(UScities_MEX, caption = "Furthest US Cities From Mexican Border",
      col.names = c("City", "State", "Distance to Mexico Border: km"),
      format.args = list(big.mark = ","))
```

2.4 - Distance to Canada (km)
```{r}
CAN=boundaries %>% 
  filter(sovereignt=='Canada')

cities = cities %>% 
  mutate(CAN_dist = st_distance(cities, CAN),
         CAN_dist = units::set_units(CAN_dist, "km"),
         CAN_dist = units::drop_units(CAN_dist))

#furthest US cities from CAN
UScities_CAN=cities %>% 
  slice_max(CAN_dist, n=5) %>% 
  select(city,state_name, CAN_dist) %>% 
  st_drop_geometry()

kable(UScities_CAN, caption = "Furthest US Cities From Canadian Border",
      col.names = c("City", "State", "Distance to Canada Border: km"),
      format.args = list(big.mark = ","))
```


3.1 Data
```{r}
#10 most populated cities in US
big_cities = cities %>%
  slice_max(population, n = 10)

ggplot()+
  geom_sf(data = boundaries)+
  geom_sf(data = USA_c)+
  geom_sf(data = big_cities, col = "red", size = 1)+
  ggthemes::theme_map()+
  labs(title = "10 Most Populated US Cities")+
  ggrepel::geom_label_repel(
    data = big_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 4
  )
```

3.2 City Distance from the Border
```{r}
farthest_2 = cities %>% 
  slice_max(dist_to_border, n = 5) %>% 
  select(city, state_name, dist_to_border)

ggplot()+
  geom_sf(data = cities, aes(col = dist_to_border), size = 1)+
  geom_sf(data = farthest_2, col = "darkblue")+
  geom_sf(data = USA_u)+
  scale_color_gradient(low = "grey", high = "red")+
  ggthemes::theme_map()+
  labs(title = "Furthest US Cities from Domestic Border")+
  ggrepel::geom_label_repel(
    data = farthest_2,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 4
  )
```
3.3 City Distance from Nearest State
```{r}
farthest_3=cities %>% 
  slice_max(dist_to_state, n = 5) %>% 
  select(city, state_name, dist_to_state)

ggplot()+
  geom_sf(data = cities, aes(col = dist_to_state), size = 1)+
  geom_sf(data = farthest_3, col = "darkblue")+
  geom_sf(data = USA_c)+
  scale_color_gradient(low = "white", high = "red")+
  ggthemes::theme_map()+
  labs(title = "City Distance from Nearest State")+
  ggrepel::geom_label_repel(
    data = farthest_3,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 4
  )
```
3.4 Equidistance boundary from Mexico and Canada
```{r}
cities = cities %>% 
  mutate(Can_Mex = abs(CAN_dist - MEX_dist))

equid_Can_Mex = cities %>% 
  filter(Can_Mex < 100)

max_equid_Can_Mex = equid_Can_Mex %>% 
  slice_max(population, n = 5)

ggplot()+
  geom_sf(data = boundaries)+
  geom_sf(data = USA_u)+
  geom_sf(data = equid_Can_Mex, color = "lightgreen", size = .5)+
  geom_sf(data = max_equid_Can_Mex, color = "red", size = .5)+
  ggthemes::theme_map()+
  labs(title = "Cities = distance from CAN&MEX border ± 100 km")+
  ggrepel::geom_label_repel(
    data = max_equid_Can_Mex,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 4
  )
```
Question 4 
```{r}
border_zone=cities %>% 
  filter(dist_to_border <= 160)
```
There are 3,296 cities are within the 100 mile border zone, or "external boundary". 


Border Zone/"External Boundary" Population is: 47,082,122
```{r}
sum(border_zone$population)
```

The % of the total population that lives in this zone is %12
```{r}
sum(border_zone$population)/sum(cities$population)
```
